Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\\: United States 2003-2012

### 2017-09-13 draft/ms-201705-draft.Rmd
updates

---
title: Confirmatory factor analysis (CFA) of multivariate lipid distributions across
  time and race-ethnic groups -- United States, 2003 to 2012
author: "Ann Von Holle, Kari E. North"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_document:
    css: style.css
    depth: 3
    highlight: tango
    number_sections: yes
    theme: readable
    toc: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
csl: ama.csl
bibliography: bib1.bib
---

@@

To assess the multivariate construct, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG. In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of race/ethnicity and sex in a United States representative sample of children ages 9 to 12 years.

@@

The primary aim is to create a latent construct of dyslipidemia. We conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG. We specified three factor loadings, one for each indicator, and 3 error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015]. All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information. We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit. BMI and age are associated with lipid levels and these two factors were allowed to covary with the latent lipid construct to improve model fit.

@@

The purpose of this paper was to examine the use of CFA as a measure of dyslipidemia. First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG. Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- common ground for comparisons. Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept. MI did not hold for different racial/ethnic nor to age groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.

@@

Testing of measurement invariance may have intriguing implications. First, there is no evidence for differences in factor loadings across any subgroups of for race/ethnicity, gender and age. This finding supports the idea of a universal system in which the three lipid indicators interact no matter the choice of sociodemographic subgroup. Confounding may play less of a part in this explanation than that for the adult population as exposure to smoking and lifestyle choices has had less time to exact a cumulative toll. Also, a similarity of factor loadings across subgroups may have relevance for lipid ratio measures such as HDL to TG, a commonly used metric. For example, a dyslipidemia factor with positive HDL-C factor loadings combined with negative LDL-C and TG factor loadings translates to a simultaneous increase of HDL and decrease in LDL-C and TG. In turn this factor interpretation could be an analog to HDL-C to TG ratio changes, implying  generalizability of these ratios across sociodemographic subgroups. No research has been done comparing these measures and would be of interest.

@@

Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research. Lipids do not function independently of one another and a measure addressing their covariance is an asset. CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators. Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.

>>>>

---
title: "Confirmatory factor analysis (CFA) of multivariate lipid distributions across time and race-ethnic groups\\: United States 2003-2012"
author: "Ann Von Holle, Kari E. North"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  word_document:
    toc: yes
  html_document:
    csl: ama.csl
    css: style.css
    depth: 3
    highlight: tango
    number_sections: yes
    theme: readable
    toc: yes
csl: ama.csl
bibliography: bib1.bib
---

@@

To assess the multivariate construct, we will use CFA to form a latent factor with three commonly assessed lipids and lipoproteins including LDL-C, HDL, TG. In turn, we will assess measurement invariance (MI)  of the construct across demographic characteristics of age, race/ethnicity and sex in a United States representative sample of children ages 12 to 19 years.

@@

The primary aim is to create a latent construct of dyslipidemia. We conducted a CFA by specifying one latent variable, dyslipidemia, with three indicators: LDL-C, HDL-C and TG. We specified three factor loadings, one for each indicator, and 3 error variances creating a ‘just identified’ model with zero degrees of freedom [@brown_confirmatory_2015]. All variables required for complex survey sampling were used in the analyses, including the sampling weights, strata and clustering information. We used the Comparative Fit Index, (CFI), adjusted Bayesian Information Criterion (aBIC) and Root Mean Square Error of Approximation (RMSEA) to select the model with best fit.

@@

The purpose of this paper was to examine the use of CFA as a measure of dyslipidemia. First, we evaluated measures of model fit of a CFA with three common lipid measures serving as observed indicators in the model: HDL-C, LDL-C and TG. Nested within this aim is the concept that this CFA dyslipidemia construct may not be similar across sociodemographic groups such as race/ethnicity and age -- common ground for comparisons. Given known differences in the factor across these demographic categories, we tested this latter hypothesis through the measurement invariance (MI) concept. MI did not hold for different racial/ethnic, age or gender groups indicating that the factor means can not be meaningfully compared across these groups, which has implications for future use.

@@

Testing of measurement invariance may have intriguing implications. First, there is no evidence for differences in factor loadings across any subgroups of for race/ethnicity, gender and age. This finding supports the idea of a universal system in which the three lipid indicators covary similarly no matter the choice of sociodemographic subgroup. Confounding may play less of a part in this explanation than that for the adult population as exposure to smoking and lifestyle choices has had less time to exact a cumulative toll. Also, a similarity of factor loadings across subgroups may have relevance for lipid ratio measures such as HDL to TG, a commonly used metric. For example, a dyslipidemia factor with positive HDL-C factor loadings combined with negative LDL-C and TG factor loadings translates to a simultaneous increase of HDL and decrease in LDL-C and TG. In turn this factor interpretation could be an analog to HDL-C to TG ratio changes, implying  generalizability of these ratios across sociodemographic subgroups. No research has been done comparing these measures and would be of interest.

@@

Measuring dyslipidemia as one joint distribution that incorporates the correlation between each of three commonly measured lipids is worthwhile either as an outcome or exposure in research. In a biologic sense, lipids do not function independently of one another and an analytic measure addressing their covariance is an asset. CFA leverages this characteristic of these measures to form a measure with different meaning from univariate indicators. Combining the observed correlations between the three common lipids alongside their common presentation in the literature and theoretical basis for dyslipidemia further justifies the use of CFA.
